Gravitational-wave sensitivity curves

### 2014-07-07 Detectors.tex
Premerge
PSD derivation

z(t,\hat{\Omega})= \frac{1}{2} \frac{\hat{p}^{j}\hat{p}^{i}} {1+\hat{\Omega}\cdot\hat{p}}\left[h_{ij}^{\mathrm{Pulsar}}\left(t-\frac{L}/c,\hat{\Omega} \right)-h_{ij}^{\mathrm{Earth}}(t,\hat{\Omega} )\right] = \frac{1}{2}\frac{\hat{p}^{j}\hat{p}^{i}}{1+\hat{\Omega}\cdot\hat{p}} \Delta h_{ij}(t,\hat{\Omega})\; .

@@

h_{ij}(t,\vec{x}) = \sum_{A\,=\,+,\,\times}\int\mathrm{d}f\;\iint_{\mathbb{S}_{2}}\mathrm{d}\hat{\Omega}\; \tilde{h}_{A}(f,\hat{\Omega})e_{ij}^{A}(\hat{\Omega}) \exp\left[2\pi i f (t-\hat{\Omega}\cdot\vec{x}\right] \; ,

@@

where $e^{A}_{ij}(\hat{\Omega})$ is the $A$ polarisation basis tensor for direction $\hat{\Omega}$  and $\vec{x}$ is the spatial position. Choosing the SSB as the origin of our coordinate system, so the pulsar is at position $L\hat{p}$, gives

@@

\left<h_{A}^{*}(f,\hat{\Omega})h_{A'}(f',\hat{\Omega}')\right> = \frac{1}{2}S_{h}(f)\delta^{(2)}(\hat{\Omega}, \hat{\Omega}')\delta_{AA'}\delta(f-f') \; ,

@@

where $\delta^{(2)}(\hat{\Omega},\hat{\Omega}')$ is the delta-function on the sphere. From (\ref{eq:45}) and (\ref{eq:stochback}), the expectation of the product of signals from two different pulsars in directions $\hat{p}_{1}$ and $\hat{p}_{2}$ may be evaluated AS

@@

\left<z_{1}(f)z_{2}^{*}(f')\right> = \frac{1}{2}S_{h}(f)\delta (f-f')\Gamma(f)

>>>>

z(t,\hat{\Omega})= \frac{1}{2} \frac{\hat{p}^{j}\hat{p}^{i}} {1+\hat{\Omega}\cdot\hat{p}}\left[h_{ij}^{\mathrm{Pulsar}}\left(t-\frac{L}{c},\hat{\Omega} \right)-h_{ij}^{\mathrm{Earth}}(t,\hat{\Omega} )\right] = \frac{1}{2}\frac{\hat{p}^{j}\hat{p}^{i}}{1+\hat{\Omega}\cdot\hat{p}} \Delta h_{ij}(t,\hat{\Omega})\; .

@@

h_{ij}(t,\vec{r}) = \sum_{A\,=\,+,\,\times}\int\mathrm{d}f\;\iint_{\mathbb{S}_{2}}\mathrm{d}\hat{\Omega}\; \tilde{h}_{A}(f,\hat{\Omega})e_{ij}^{A}(\hat{\Omega}) \exp\left[2\pi i f \left(t-\frac{\hat{\Omega}\cdot\vec{x}}{c}\right)\right] \; ,

@@

where $e^{A}_{ij}(\hat{\Omega})$ is the $A$ polarisation basis tensor for direction $\hat{\Omega}$  and $\vec{r}$ is the spatial position. Choosing the SSB as the origin of our coordinate system, so the pulsar is at position $L\hat{p}$, gives

@@

\left<\tilde{h}_{A}^{*}(f,\hat{\Omega})\tilde{h}_{A'}(f',\hat{\Omega}')\right> = \frac{1}{2}S_{h}(f)\delta^{(2)}(\hat{\Omega}, \hat{\Omega}')\delta_{AA'}\delta(f-f') \; ,

@@

where $\delta^{(2)}(\hat{\Omega},\hat{\Omega}')$ is the delta-function on the sphere. From (\ref{eq:45}) and (\ref{eq:stochback}), the expectation of the product of signals from two different pulsars in directions $\hat{p}_{1}$ and $\hat{p}_{2}$ may be evaluated as

@@

\left<\tilde{z}_{1}(f)\tilde{z}_{2}^{*}(f')\right> = \frac{1}{2}S_{h}(f)\delta (f-f')\Gamma(f)

### 2014-07-07 Introduction.tex
PSD derivation

Gravitational radiation has two independent polarization states denoted $+$ and $\times$; a general signal can be described as a linear combination of the two polarization states, $h \nobreak=\nobreak A_{+} h_{+}\nobreak +\nobreak A_{\times} h_{\times}$. The sensitivity of a detector to these depends upon the relative orientations of the source and detector. The output of a gravitational wave detector $s(t)$ contains a superposition of noise $n(t)$ and (possibly) the signal $h(t)$,
\begin{equation}
s(t) = n(t)+h(t) \; .
\end{equation}
We will work with the Fourier transform of the signal, using the conventions that
\begin{eqnarray} \label{eq:defFourier}
\tilde{x}(f) &= {\mathcal{F}}\left\{x(t)\right\} &= \int_{-\infty}^{\infty}\mathrm{d}t\;x(t)\exp (-2\pi \rmi f t ) \; , \\
x(t) &= {\mathcal{F}}^{-1}\left\{\tilde{x}(f)\right\} &= \int_{-\infty}^{\infty}\mathrm{d}f\;\tilde{x}(f)\exp (2\pi \rmi f t ) \; .
\end{eqnarray}

@@

When integrated over all positive frequencies, the PSD gives the mean square noise amplitude. Starting with taking the \emph{time} average of the square of the detector noise:
\begin{eqnarray}\label{eq:meansquare}
\overline{n(t)^{2}} &= \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T}\mathrm{d}t\; n(t) ^{2} \nonumber \\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T}\mathrm{d}t\; n(t)n^{*}(t) \nonumber \\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\tilde{n}(f)\tilde{n}^{*}(f')e^{2\pi \rmi ft }e^{-2\pi \rmi f't } \nonumber\\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\cal{F}}\left\{ n(\tau) \right\}(f){\cal{F}}\left\{ n(\tau) \right\}(f')^{*}e^{2\pi \rmi ft }e^{-2\pi \rmi f't }
\end{eqnarray}
Where in the penultimate line we have used the definition of the Fourier transform, (\ref{eq:defFourier}), and in the last line we have just changed notation. A property of Fourier transforms is that in the time domain a translation by amount $t$ translates to a rotation in phase by $2\pi ft$ in the frequency domain; if ${\mathcal{F}}\left\{ n(\tau) \right\} = \tilde{n}(f)$ then ${\mathcal{F}}\left\{ n(\tau-t) \right\} = \tilde{n}(f)\exp (2\pi \rmi ft )$. Therefore the exponential factors in (\ref{eq:meansquare}) may be absorbed as a time-translation of the noise by an amount $t$
\begin{eqnarray}
\overline{n(t)^{2}} & = \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T}\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\cal{F}}\left\{ n(\tau-t) \right\}(f){\cal{F}}\left\{ n(\tau-t) \right\}(f')^{*} \, .
\end{eqnarray}
Using the ergodic property of the system described above the time average (denoted by $\overline{(\ldots)}$) may now be changed into an ensemble average (denoted by $\left<\ldots\right>$).
The noise is stationary so it's expectation value is unchanged by the time translation performed above. Therefore using (\ref{eq:psd}) gives
\begin{eqnarray}\label{eq:meansquare1}
&=\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>\nonumber \\
&=\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\frac{1}{2}S_{n}(f)\delta(f-f') \nonumber \\
& = \int_{0}^{\infty}\mathrm{d}f\; S_{n}(f)
\end{eqnarray}

Given a detector output, the challenge is to extract the signal. There is a well known solution to this problem which involves constructing a Wiener optimal filter \citep{Wiener49}. Let $K(t)$ be the filter function with Fourier transform $\tilde{K}(f)$. Convolving this with the detector output gives a contribution from the signal and a contribution from the noise
\begin{equation}\label{eq:conv}
\left(s*K\right)(\tau) = \int_{-\infty}^{\infty}\mathrm{d}t\;(n(t)+h(t))K(t-\tau) \approx \mathcal{S} + \mathcal{N} \; .
\end{equation}
The signal contribution $\mathcal{S}$ is defined as the expectation of the convolution in (\ref{eq:conv}) when a signal is present, maximised by varying the offset $\tau$. \cplb{I think this needs expanding.} Since the expectation of pure noise is zero it, follows that
\begin{equation}
\mathcal{S} = \int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K(t)=\int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K^{*}(t)=\int_{-\infty}^{\infty}\mathrm{d}f\; \tilde{h}(f)\tilde{K}^{*}(f) \; .
\end{equation}
The squared contribution from noise $\mathcal{N}^{2}$ is defined as the mean square of the convolution in (\ref{eq:conv}) when no signal is present,
\begin{eqnarray} 
\mathcal{N}^{2} &= \int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K(t')\left<n(t)n(t')\right> \nonumber \\
 &= \int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K^{*}(t')\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>\exp\left[2\pi\rmi(ft-f't')\right]\nonumber \\
%&=\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;\frac{1}{2}S_{n}(f)K(t)K^{*}(t')\exp\left(2\pi\rmi f (t-t')\right)\nonumber \\
 &= \int_{-\infty}^{\infty}\mathrm{d}f \; \frac{1}{2}S_{n}(f)\tilde{K}(f)\tilde{K}^{*}(f) \; ,
 \end{eqnarray}
using the definition of $S_{n}(f)$ from (\ref{eq:psd}). Hence the signal-to-noise ratio (SNR) $\varrho$ is given by
\begin{equation}\label{eq:SNRinnerprod} 
\varrho^{2} = \frac{\mathcal{S}^{2}}{\mathcal{N}^{2}}= \frac{\innerprod{\frac{1}{2}S_{n}(f)\tilde{K}(f)}{\tilde{h}}^{2}}{\innerprod{\frac{1}{2}S_{n}(f)\tilde{K}(f)}{\frac{1}{2}S_{n}(f)\tilde{K}(f)}} \;,
\end{equation}
where we have introduced the inner product between signal $A$ and $B$ as \citep{Finn1992}
\begin{equation}\label{eq:snrinnerprod} \left( \tilde{A}(f) | \tilde{B}(f) \right) = 4\Re\left\{\int_{0}^{\infty}\mathrm{d}f\;\frac{\tilde{A}^{*}(f)\tilde{B}(f)}{S_{n}(f)}\right\} \; .\end{equation}
The optimum filter maximises the SNR in (\ref{eq:SNRinnerprod}). From the Cauchy--Schwarz inequality, it follows that the optimum filter is
\begin{equation}
\tilde{K}(f)=\frac{\tilde{h}(f)}{S_{n}(f)} \; .
\end{equation}
This is the Wiener filter, it may be multiplied by an arbitrary constant since this does not change the SNR. Using this form for $\tilde{K}(f)$, the squared SNR is
\begin{equation}
\varrho^2 = \int_0^\infty\mathrm{d}f \frac{4| \tilde{h}(f)|^{2}}{S_n(f)} = \innerprod{h}{h}.
\label{eq:traditionalSNR} 
\end{equation}
In order to construct the Wiener filter, it is necessary to know \emph{a priori} the form of the signal, $\tilde{h}(f)$.

@@

Cosmological studies often work in terms of the dimensionless quantity $\Omega_{\mathrm{GW}}$, the energy density per logarithmic frequency interval normalised to the critical density of the universe:
\begin{equation}
\label{eq:omega}
\Omega_\mathrm{GW}(f) = \frac{fS_{\mathrm{E}}(f)}{\rho_{\mathrm{c}}c^{2}} \; .
\end{equation}
The critical density of the universe is
\begin{equation}\label{eq:crit-density}
\rho_{\mathrm{c}}=3H_{0}^{2}/8\pi G \;,
\end{equation}
 where $H_{0}$ is the Hubble constant which is commonly parametrized as
\begin{equation}
H_0 =h_{100}\times 100~\mathrm{km\,s^{-1}\,Mpc^{-1}}.
\end{equation}
The reduced Hubble parameter $h_{100}$ has nothing to do with strain. The most common quantity related to energy density to be plotted on sensitivity curves is $\Omega_{\mathrm{GW}}h_{100}^{2}$ (figure \ref{fig:omega}) as this removes sensitivity to the measured value of the Hubble constant.

>>>>

Gravitational radiation has two independent polarization states denoted $+$ and $\times$; a general signal can be described as a linear combination of the two polarization states, $h \nobreak=\nobreak A_{+} h_{+}\nobreak +\nobreak A_{\times} h_{\times}$. The sensitivity of a detector to these depends upon the relative orientations of the source and detector. The output of a gravitational wave detector $s(t)$ contains a superposition of noise $n(t)$ and (possibly) the signal $h(t)$,
\begin{equation}
s(t) = n(t)+h(t) \; .
\end{equation}
We shall work with the Fourier transform of the signal, using the conventions that
\begin{eqnarray} \label{eq:defFourier}
\tilde{x}(f) &= {\mathcal{F}}\left\{x(t)\right\}(f) &= \int_{-\infty}^{\infty}\mathrm{d}t\;x(t)\exp (-2\pi \rmi f t ) \; , \\
x(t) &= {\mathcal{F}}^{-1}\left\{\tilde{x}(f)\right\}(t) &= \int_{-\infty}^{\infty}\mathrm{d}f\;\tilde{x}(f)\exp (2\pi \rmi f t ) \; .
\end{eqnarray}

@@

When integrated over all positive frequencies, the PSD gives the mean square noise amplitude. Starting by taking the time average of the square of the detector noise:
\begin{eqnarray}\label{eq:meansquare}
\overline{\left|n(t)\right|^{2}} &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; n(t)n^{*}(t) \\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\tilde{n}(f)\tilde{n}^{*}(f')\exp\left(2\pi \rmi ft\right)\exp\left(-2\pi \rmi f't\right) \nonumber\\
 &= \lim_{T\rightarrow\infty}\frac{1}{2T} & \int_{-T}^{T}\mathrm{d}t\; \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\mathcal{F}}\left\{ n(\tau) \right\}(f)\left[{\mathcal{F}}\left\{ n(\tau) \right\}(f')\right]^{*} \nonumber \\*
 & & \times \exp\left(2\pi \rmi ft\right)\exp\left(-2\pi \rmi f't\right)
\end{eqnarray}
Where we have substituted in using the definitions of the Fourier transform and its inverse. A property of Fourier transforms is that in the a time-domain translation by amount $t$ translates to a frequency-domain phase change $2\pi ft$; if ${\mathcal{F}}\left\{ n(\tau) \right\}(f) = \tilde{n}(f)$, then ${\mathcal{F}}\left\{ n(\tau-t) \right\}(f) = \tilde{n}(f)\exp (2\pi \rmi ft )$. Therefore, the exponential factors in (\ref{eq:meansquare}) may be absorbed as
\begin{eqnarray}
\overline{\left|n(t)\right|^{2}} &=& \lim_{T\rightarrow\infty}\frac{1}{2T}\int_{-T}^{T} \mathrm{d}t\:\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;{\mathcal{F}}\left\{ n(\tau-t) \right\}(f)\left[{\mathcal{F}}\left\{ n(\tau-t) \right\}(f')\right]^{*} \, .
\end{eqnarray}
Since the noise is a randomly varying signal, we can use the ergodic principle to equate a time average, denoted by $\overline{(\ldots)}$, with an ensemble average, denoted by $\left<\ldots\right>$. The noise is stationary, consequently, its expectation value is unchanged by time-translation performed above. Therefore,
\begin{eqnarray}\label{eq:meansquare1}
\overline{\left|n(t)\right|^{2}} &=& \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right> \\
&=& \int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\frac{1}{2}S_{n}(f)\delta(f-f') \nonumber \\
&=& \int_{0}^{\infty}\mathrm{d}f\; S_{n}(f),
\end{eqnarray}
using (\ref{eq:psd}).

Given a detector output, the challenge is to extract the signal. There is a well known solution to this problem which involves constructing a Wiener optimal filter \citep{Wiener49}. Let $K(t)$ be the filter function with Fourier transform $\tilde{K}(f)$. Convolving this with the detector output gives a contribution from the signal and a contribution from the noise
\begin{equation}\label{eq:conv}
\left(s*K\right)(\tau) = \int_{-\infty}^{\infty}\mathrm{d}t\;\left[h(t)+n(t)\right]K(t-\tau) \approx \mathcal{S} + \mathcal{N} \; .
\end{equation}
The signal contribution $\mathcal{S}$ is defined as the expectation of the convolution in (\ref{eq:conv}) when a signal is present, maximised by varying the offset $\tau$. \cplb{I think this needs expanding.} Since the expectation of pure noise is zero it, follows that
\begin{equation}
\mathcal{S} = \int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K(t)=\int_{-\infty}^{\infty}\mathrm{d}t\;h(t)K^{*}(t)=\int_{-\infty}^{\infty}\mathrm{d}f\; \tilde{h}(f)\tilde{K}^{*}(f) \; .
\end{equation}
The squared contribution from noise $\mathcal{N}^{2}$ is defined as the mean square of the convolution in (\ref{eq:conv}) when no signal is present,
\begin{eqnarray} 
\mathcal{N}^{2} &= \int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K(t')\left<n(t)n(t')\right> \nonumber \\
 &= \int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;K(t)K^{*}(t')\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}f'\;\left<\tilde{n}(f)\tilde{n}^{*}(f')\right>\exp\left[2\pi\rmi(ft-f't')\right]\nonumber \\
%&=\int_{-\infty}^{\infty}\mathrm{d}f\;\int_{-\infty}^{\infty}\mathrm{d}t\;\int_{-\infty}^{\infty}\mathrm{d}t'\;\frac{1}{2}S_{n}(f)K(t)K^{*}(t')\exp\left(2\pi\rmi f (t-t')\right)\nonumber \\
 &= \int_{-\infty}^{\infty}\mathrm{d}f \; \frac{1}{2}S_{n}(f)\tilde{K}(f)\tilde{K}^{*}(f) \; ,
 \end{eqnarray}
using the definition of $S_{n}(f)$ from (\ref{eq:psd}). Hence the signal-to-noise ratio (SNR) $\varrho$ is given by
\begin{equation}\label{eq:SNRinnerprod} 
\varrho^{2} = \frac{\mathcal{S}^{2}}{\mathcal{N}^{2}}= \frac{\innerprod{\frac{1}{2}S_{n}(f)\tilde{K}(f)}{\tilde{h}(f)}^{2}}{\innerprod{\frac{1}{2}S_{n}(f)\tilde{K}(f)}{\frac{1}{2}S_{n}(f)\tilde{K}(f)}} \;,
\end{equation}
where we have introduced the inner product between signal $\tilde{A}$ and $\tilde{B}$ as \citep{Finn1992}
\begin{equation}\label{eq:snrinnerprod} \innerprod{\tilde{A}(f)}{\tilde{B}(f)} = 4\Re\left\{\int_{0}^{\infty}\mathrm{d}f\;\frac{\tilde{A}^{*}(f)\tilde{B}(f)}{S_{n}(f)}\right\} \; .\end{equation}
The optimum filter maximises the SNR in (\ref{eq:SNRinnerprod}). From the Cauchy--Schwarz inequality, it follows that the optimum filter is
\begin{equation}
\tilde{K}(f)=\frac{\tilde{h}(f)}{S_{n}(f)} \; .
\end{equation}
This is the Wiener filter, it may be multiplied by an arbitrary constant since this does not change the SNR. Using this form for $\tilde{K}(f)$, the squared SNR is
\begin{equation}
\varrho^2 = \int_0^\infty\mathrm{d}f \frac{4| \tilde{h}(f)|^{2}}{S_n(f)} = \innerprod{\tilde{h}(f)}{\tilde{h}(f)}.
\label{eq:traditionalSNR} 
\end{equation}
In order to construct the Wiener filter, it is necessary to know \emph{a priori} the form of the signal, $\tilde{h}(f)$.

@@

Cosmological studies often work in terms of the dimensionless quantity $\Omega_{\mathrm{GW}}$, the energy density per logarithmic frequency interval normalised to the critical density of the universe:
\begin{equation}
\label{eq:omega}
\Omega_\mathrm{GW}(f) = \frac{fS_{\mathrm{E}}(f)}{\rho_{\mathrm{c}}c^{2}} \; .
\end{equation}
The critical density of the universe is
\begin{equation}\label{eq:crit-density}
\rho_{\mathrm{c}}=\frac{3H_{0}^{2}}{8\pi G} \;,
\end{equation}
 where $H_{0}$ is the Hubble constant which is commonly parametrized as
\begin{equation}
H_0 =h_{100}\times 100~\mathrm{km\,s^{-1}\,Mpc^{-1}}.
\end{equation}
The reduced Hubble parameter $h_{100}$ has nothing to do with strain. The most common quantity related to energy density to be plotted on sensitivity curves is $\Omega_{\mathrm{GW}}h_{100}^{2}$ (figure \ref{fig:omega}) as this removes sensitivity to the measured value of the Hubble constant.
